Plume-SPH 1.0: A 3D dust-gas volcano plume model based on smoothed particle hydrodynamics

### 2016-10-18 Manuscripts_GMD.tex
update1
updated according to Dr. Bursik's comments

Volcanic ash, if persists in the atmosphere, will pose great threat on aircrafts and communities. VATDs (Volcanic ash transport and dispersion models) usually take output of plume models as source terms. The accuracy of source terms is crucial for VATDs. All of existing 3D (three dimensional) plume models are using mesh based methods. SPH (Smoothed particle hydrodynamics), as a mesh free method has several advantages over mesh based methods in modeling of multiphase free boundary flow. As an initial effort on exploiting the feasibility and advantage of SPH in volcano plume modeling we adopt a relative simpler model, a 3D dusty-gas dynamic model, targeting at capturing the more basic features of volcano plume. In this model, erupted material is assumed to be well mixed and represented by one phase while air is another phase. Immediate dynamic and thermal dynamic equilibrium between air and erupted material are assumed. Effect of wind is not considered. Several newly developed techniques in SPH are adopted and trimmed to address numerical challenges in simulating multiphase compressible turbulent flow with classical SPH method. The $SPH-\varepsilon$ turbulence model is adopted to capture sub-particle mixing. Heat exchange due to turbulence is then calculated with Reynolds analogy. Corrected formulism of SPH is used to handle tensile instability and deficiency of particle distribution near the boundaries. We also propose feasible ways to impose velocity inlet and pressure outlet boundary conditions, both of which are scarce in traditional implementations of SPH. The core solver of our model is parallelized with MPI (message passing interface) obtaining good weak scalability and strong scalability. The model is first verified by comparing velocity and concentration distribution along axis and cross section with experimental results of JPUE (jet or plume which is ejected from a nozzle into a uniform environment). The top height of Pinatubo eruption (15 June 1991) simulated by our model is consistent with both observation and existing 3D plume models. Profiles of several integrated variables are compared with these by existing 3D plum models and further verifies the correctness of our model. Analysis on the plume evolution process illustrated that this model is able to reproduce the basic physics of plume development. The comparison also implies that turbulence model plays a significant role in 3D volcano plume modeling.

@@

The dynamics of explosive eruption clouds has been the central issue of volcanology science for a long time. Primary hazards associated with explosive grey volcanoes include pyroclastic density currents (flows and surges), the widespread deposition of airfall tephra, and the threats posed by volcanic ash. In addition, there are also other secondary hazards associated with volcanoes \citep{lockwood2013volcanoes}.

@@

The basic physics of volcanic plume has been understood for more than half a century \citep{morton1956turbulent, settle1978volcanic, wilson1978control}. Several volcano plume models have been developed in the past few decades. Specific aspects of a plume have been studied by 1D (one dimensional) plume models which describe the steady state solution of a plume under idealized boundary conditions.
The effects of magma type and vent conditions \citep{woods1988fluid, woods1995decompression}, atmospheric conditions \citep{ woods1993moist, sparks1997volcanic, bursik2001effect}, external surface water \citep{koyaguchi1996formation}, thermal disequilibrium, and particle fallout \citep{woods1991particle} have been assessed in increasing details through 1D numerical simulations. Recently developed 1D models (eg. \citep{mastin2007user, woodhouse2013interaction}) tend to account for more physics effects in one model. FPLUME-1.0 \citep{folch2016fplume} accounts for wind effect, entrainment of moisture, water phase change, particle fallout and re-entrainment and even wet aggregation of ash. However, In these 1D models, the entrainment of air is evaluated based on two coefficients: entrainment coefficient for the vertical plume and the entrainment coefficient that describes the effect of wind. Different 1D models adopt different entrainment coefficients based on specific formulation or calibration against well-documented case studies. The feedback from plume to atmosphere is ignored in 1D models. Even though determination of essential parameters such as the entrainment is not based on first principles, such simple models nevertheless allow us to investigate the importance of physical mechanisms in a volcanic plume. In addition, these simplified models requires little computational resource and can run on standard personal computers or on web sites in very short time. As a result, 1D software for volcano plume development (such as \citep{267, 1194, 3541}), combined with VATDs (such as \citep{114, draxler2015hysplit}) are widely used in research and practice. While these 1D models can generate well-matched results with 3D (three dimensional) models for weak plumes, much greater variability is observed for strong plume scenarios, especially for local variables \citep{costa2016results}. In addition, the need for greater skills in associated hazards forecasts with 1D models requires 2D (two dimensional) and 3D models.\\

@@

Another 3D model, SK-3D \citep{suzuki2005numerical} is a three-dimensional time-dependent fluid dynamics model that can reproduce the entrainment process of eruption clouds with relative simpler physics but high order of accuracy and high spatial resolution, which is applicable to time-dependent phenomena in actual volcanological simulations. 

@@

All of these existing 3D models are based on mesh based methods (Eulerian methods). And there is no 3D plume model based on mesh free method (Lagrange methods) yet. Actually, Lagrangian methods have several features that are suitable for volcano plume simulation. Among such Lagrangian methods, Smoothed Particle Hydrodynamics \citep{gingold1977smoothed,lucy1977numerical}, even though relatively newer compared to well designed mesh based methods such as Finite Difference, has shown good agreement with experiments by many applications in fluid dynamic area. And it currently is, by far the most widely used scheme. Specifically, the reason we choose SPH as the numerical method for volcanic plume simulation are as following:

@@

@@

\label{fig:strong_elevation}

@@

%\begin{figure}
%\center
%\includegraphics[width=5cm]{t15}
%\caption{Contour of mass fraction and velocity quiver at 15 seconds, column rises up quickly due to initial momentum.}
%\label{fig:strong_t15}
%\end{figure}

@@

Figure \ref{fig:strong_elevation} shows the overall evolution process of plume. Plume keeps rising up and entraining air into the plume until it reaches its top height(around t=300 seconds). Afterwards the mixtures of erupted material and air then fall back to neutral height and starts spreading radially. Detailed velocity quiver plot in Fig. \ref{fig:strong_plume_t50} shows entraining of air into the plume at the edge of column. The evolution of the plume further verifies our model.

@@

A new plume model was developed based on SPH method. Advanced numerical techniques in SPH were exploited and tailored for this model. High performance computing was used to mitigate confliction between accuracy (depends on comprehensiveness of the model, resolution, order of accuracy of numerical methods, scheme for time upgrading.) and simulation time (depends on, comprehensiveness, resolution, order of accuracy of numerical methods, scheme for time upgrading ... and computational techniques). The correctness of the code and model was verified and validated by a series of test simulations.\\

>>>>

Volcanic ash, if persists in the atmosphere, will pose great threat on aircrafts and communities. VATDs (Volcanic ash transport and dispersion models) usually take output of plume models as source terms. The accuracy of source terms is crucial for VATDs. All of existing 3D (three dimensional) plume models are using mesh based methods. SPH (Smoothed particle hydrodynamics), as a mesh free method has several advantages over mesh based methods in modeling of multiphase free boundary flow. As an initial effort on exploiting the feasibility and advantage of SPH in volcano plume modeling we adopt a relative simpler model, a 3D dusty-gas dynamic model, targeting at capturing the more basic features of volcano plume. In this model, erupted material is assumed to be well mixed and represented by one phase while air is another phase. Immediate dynamic and thermal dynamic equilibrium between air and erupted material are assumed. Effect of wind is not considered. Several newly developed techniques in SPH are adopted and trimmed to address numerical challenges in simulating multiphase compressible turbulent flow with classical SPH method. The $SPH-\varepsilon$ turbulence model is adopted to capture sub-particle mixing. Heat exchange due to turbulence is then calculated with Reynolds analogy. Corrected formulism of SPH is used to handle tensile instability and deficiency of particle distribution near the boundaries. We also propose feasible ways to impose velocity inlet and pressure outlet boundary conditions, both of which are scarce in traditional implementations of SPH. The core solver of our model is parallelized with MPI (message passing interface) obtaining good weak scalability and strong scalability. The model is first verified by comparing velocity and concentration distribution along axis and cross section with experimental results of JPUE (jet or plume which is ejected from a nozzle into a uniform environment). The top height of Pinatubo eruption (15 June 1991) simulated by our model is consistent with both observation and existing 3D plume models. Profiles of several integrated variables are compared with these by existing 3D plume models and further verifies the correctness of our model. Analysis on the plume evolution process illustrated that this model is able to reproduce the basic physics of plume development. The comparison also implies that turbulence model plays a significant role in 3D volcano plume modeling.

@@

The dynamics of explosive eruption clouds has been the central issue of volcanology science for a long time \citep{settle1978volcanic, wilson1978control}.
The basic physics of volcanic plume has been understood for more than half a century \citep{morton1956turbulent, settle1978volcanic, wilson1978control}. The effects of magma type and vent conditions \citep{woods1988fluid, woods1995decompression}, atmospheric conditions \citep{ woods1993moist, sparks1997volcanic, bursik2001effect}, external surface water \citep{koyaguchi1996formation}, thermal disequilibrium, and particle fallout \citep{woods1991particle} have been assessed in increasing details based on observation, experiments and numerical studies.
Hazards associated with explosive grey volcanoes include pyroclastic density currents (flows and surges), the widespread deposition of airfall tephra, and the threats posed by volcanic ash. 
%In addition, there are also other secondary hazards associated with volcanoes \citep{lockwood2013volcanoes}.

@@

Several 1D (one dimensional) volcano plume models have been developed in the past few decades. Based on the most basic 1D model \citep{woods1988fluid} which only accounts for basis conservation, recently developed 1D models (eg. \citep{bursik2001effect, mastin2007user, degruyter2012improving, woodhouse2013interaction, devenish2013using, de2015plume, folch2016fplume, pouget2016sensitivity}) tend to account for more physics effects. FPLUME-1.0 \citep{folch2016fplume} accounts for wind effect, entrainment of moisture, water phase change, particle fallout and re-entrainment and even wet aggregation of ash. However, In these 1D models, the entrainment of air is evaluated based on two coefficients: entrainment coefficient for the vertical plume and the entrainment coefficient that describes the effect of wind. Different 1D models adopt different entrainment coefficients based on specific formulation or calibration against well-documented case studies. The feedback from plume to atmosphere is also ignored in 1D models. Even though determination of essential parameters such as the entrainment is not based on first principles, such simple models nevertheless allow us to investigate the importance of physical mechanisms in a volcanic plume. In addition, these simplified models requires little computational resource and can run on standard personal computers or on web sites in very short time. As a result, 1D software for volcano plume development (such as \citep{267, 1194, 3541}), combined with VATDs (such as \citep{114, draxler2015hysplit}) are widely used in research and practice. While these 1D models can generate well-matched results with 3D (three dimensional) models for weak plumes, much greater variability is observed for strong plume scenarios, especially for local variables \citep{costa2016results}. In addition, the need for greater skills in associated hazards forecasts with 1D models requires 2D (two dimensional) and 3D models.\\

@@

Another 3D model, SK-3D \citep{suzuki2005numerical} is a 3D time-dependent fluid dynamics model that can reproduce the entrainment process of eruption clouds with relative simpler physics but high order of accuracy and high spatial resolution, which is applicable to time-dependent phenomena in actual volcanological simulations. 

@@

All of these existing 3D models are based on mesh based methods (Eulerian methods). And there is no 3D plume model based on mesh free method (Lagrange methods) yet. Actually, Lagrangian methods have several features that are suitable for volcano plume simulation. Among such Lagrangian methods, Smoothed Particle Hydrodynamics \citep{gingold1977smoothed,lucy1977numerical}, even though relatively newer compared to well designed mesh based methods such as Finite Difference, has shown good agreement with experiments by many applications in fluid dynamic area. And it currently is, by far the most widely used scheme. Implementations of SPH in volcanology are keeping showing up \citep{bursik2003smoothed,herault2010sph,haddad2016smoothed}.\\
Specifically, the reasons that motivate us to choose SPH as the numerical method for volcanic plume simulation are as following:

@@

\item Investigation of mixing phenomena. 
\item ZFE (Zone of Flow Establishment), ZEF (Zone of Established Flow) investigation and relation to column collapse and the mystery of the development of entrainment and proper characterization through numerical experimentation.
\item Related to below, easy inclusion of particles of different sizes (phases) and investigation of detailed mechanics of sedimentation and drag force interaction in lower plume.

@@

\label{fig:strong_evolution}

@@

%\includegraphics[width=5cm]{t15}
%\caption{Contour of mass fraction and velocity quiver at 15 seconds, column rises up quickly due to initial momentum.}
%\label{fig:strong_t15}

@@

Figure \ref{fig:strong_evolution} shows the overall evolution process of plume. Plume keeps rising up and entraining air into the plume until it reaches its top height(around t=300 seconds). Afterwards the mixtures of erupted material and air then fall back to neutral height and starts spreading radially. Detailed velocity quiver plot in Fig. \ref{fig:strong_plume_t50} shows entraining of air into the plume at the edge of column. The evolution of the plume further verifies our model.

@@

A new plume model was developed based on SPH method. Advanced numerical techniques in SPH were exploited and tailored for this model. High performance computing was used to mitigate confliction between accuracy (depends on comprehensiveness of the model, resolution, order of accuracy of numerical methods, scheme for time upgrading.) and simulation time (depends on, comprehensiveness, resolution, order of accuracy of numerical methods, scheme for time upgrading ... and computational techniques). The correctness of the code and model was verified and validated by a series of test simulations. Dimensionless velocity and concentration distribution across the cross-section and along the jet axis match well with experimental results of JPUE. Top height and integrated local variables simulated by our model are consistent with both observation and simulation results of existing 3D plume models. Comparison between our results with that of SK-3D implies that turbulence model plays a significant role in plume modeling.\\
